Characterizing Signal Loss, Error, and Bias in the 21cm Reionization \\Power Spectrum: A Revised Study of PAPER-64

### 2018-05-10 PAPER.tex
listed bootstrapping changes in bullet-point form; added some text in the jackknife section about the cross-term not averaging down

We bootstrap PAPER-64 power spectra in order to determine confidence intervals for our results. In this section, we highlight 
one major change in the way we estimate errors since \citetalias{ali_et_al2015}, using the lesson we have learned about bootstrapping independent samples.

As discussed in Section \ref{sec:ErrorOverview}, bootstrapping is only a valid way of estimating errors if a dataset is comprised 
of independent samples, or the number of independent samples is well known. The PAPER-64 pipeline outputs $20$ bootstraps (over baselines), each a $2$-dimensional power 
spectrum that is a function of $k$ and time.

@@

Given our new understanding of the sensitivity of bootstraps to the number of elements sampled, we have removed the second 
bootstrapping step along time entirely and now simply bootstrap over the baseline axis. Power spectrum $2\sigma$ errors with this bootstrapping change for fringe-rate filtered noise are shown in Figure 
\ref{fig:data_errors}. The estimates are uniformly weighted in order to disentangle the effects of bootstrapping from signal loss. As 
shown in the figure, when more elements are drawn for each bootstrap than the number of 
independent samples (by over-sampling elements along the time axis), repeated values begin to crop up and the apparent variation between bootstraps drops, resulting in limits (gray) below the predicted noise level (green). Using the revised bootstrapping method, where bootstrapping only occurs over the baseline axis, the limits (black) are shown to better agree with the analytic prediction for noise. While Figure \ref{fig:data_errors} implies that errors are under-estimated by a factor of $\sim$ $5$ in mK$^{2}$ for the noise simulation, in practice this factor is lower for the case of real data (a factor of $\sim$ $2$ in mK$^{2}$ instead), possibly due to the data being less correlated in time than the fringe-rate filtered noise in the simulation. %Figure \ref{fig:data_errors} implies that power spectrum upper limits were originally under-estimated by a factor of $\sim$ $5$ in mK$^{2}$.

In addition to no longer bootstrapping along the time axis, we have made two other changes from the \citetalias{ali_et_al2015} analysis regarding bootstrapping. The first is that we now bootstrap over baseline cross-products, instead of baselines themselves. In the previous analysis, baselines were bootstrapped prior to forming cross power spectra, and using this particular ordering of operations (bootstrapping, then cross-multiplication) yields variances that have been found to disagree with predicted errors from bootstrapping using simulations. On the contrary, bootstrapping over cross power spectra ensures that we are estimating the variance of our quantity of interest (i.e. the power spectrum). This change, while fundamental in retaining the integrity of the bootstrapping method in general, alters the resulting power spectrum errors by factors of $<2$ in practice.

Finally, the last change from the \citetalias{ali_et_al2015} method is that our power spectrum points (previously computed as the mean of all bootstraps), are now computed as the power spectrum estimate resulting from not bootstrapping at all. More specifically, we compute one estimate without sampling, and this estimate is propagated through our signal loss computation (this estimate is $\widehat{\textbf{P}}_{x}$). The difference between taking the mean of the bootstrapped values and using the estimate from the no-bootstrapping case is small, but doing the latter ensures that we are forming results that reflect the estimate preferred by all our data.

@@

The second null test possibility (error bars consistent with zero but not with thermal noise) is displayed by the $k$-values just outside the horizon ($k\sim\pm 0.1$\,$h$ Mpc$^{-1}$) for all three tests. This indicates an additive noise component that is increasing our errors. More specifically, although we expect each cross-multiplication that is used in power spectrum estimation to have independent noise, there is still the possibility of a noise-foreground coupling term that can introduce power. This is because cross-multiplications produce four additive terms --- a signal-squared term (where signal includes both foregrounds and EoR), two cross-terms between the signal and noise, and one noise-only term. When differencing two power spectra (each with their own four terms), we expect the signal-only term to subtract out for a ``passing" null test, and we expect the noise-only terms to be consistent with thermal noise. The cross-terms, however, can introduce extra variance that would vary with foreground strength. It is therefore not surprising that this effect is largest at $k$-values just outside the horizon, where we expect foregrounds to be brightest post-delay filtering.

>>>>

We bootstrap PAPER-64 power spectra in order to determine confidence intervals for our results. In this section, we highlight 
four specific changes in the way we estimate errors since \citetalias{ali_et_al2015}, the first of which builds off of the lesson we have learned previously about bootstrapping independent samples.

\begin{itemize}

\item{As discussed in Section \ref{sec:ErrorOverview}, bootstrapping is only a valid way of estimating errors if a dataset is comprised 
of independent samples, or the number of independent samples is well known. The PAPER-64 pipeline outputs $20$ bootstraps (over baselines), each a $2$-dimensional power 
spectrum that is a function of $k$ and time.

@@

Given our new understanding of the sensitivity of bootstraps to the number of elements sampled, we have removed the second 
bootstrapping step along time entirely and now simply bootstrap over the baseline axis. Power spectrum $2\sigma$ errors with this bootstrapping change for fringe-rate filtered noise are shown in Figure 
\ref{fig:data_errors}. The estimates are uniformly weighted in order to disentangle the effects of bootstrapping from signal loss. As 
shown in the figure, when more elements are drawn for each bootstrap than the number of 
independent samples (by over-sampling elements along the time axis), repeated values begin to crop up and the apparent variation between bootstraps drops, resulting in limits (gray) below the predicted noise level (green). Using the revised bootstrapping method, where bootstrapping only occurs over the baseline axis, the limits (black) are shown to better agree with the analytic prediction for noise. While Figure \ref{fig:data_errors} implies that errors are under-estimated by a factor of $\sim$ $5$ in mK$^{2}$ for the noise simulation, in practice this factor is lower for the case of real data (a factor of $\sim$ $2$ in mK$^{2}$ instead), possibly due to the data being less correlated in time than the fringe-rate filtered noise in the simulation. }%Figure \ref{fig:data_errors} implies that power spectrum upper limits were originally under-estimated by a factor of $\sim$ $5$ in mK$^{2}$.

\item{A second change to our bootstrapping procedure is that we now bootstrap over baseline cross-products, instead of baselines themselves. In the previous analysis, baselines were bootstrapped prior to forming cross power spectra, and using this particular ordering of operations (bootstrapping, then cross-multiplication) yields variances that have been found to disagree with predicted errors from bootstrapping using simulations. On the contrary, bootstrapping over cross power spectra ensures that we are estimating the variance of our quantity of interest (i.e. the power spectrum). This change, while fundamental in retaining the integrity of the bootstrapping method in general, alters the resulting power spectrum errors by factors of $<2$ in practice.}

\item{In \citetalias{ali_et_al2015}, individual baselines were divided into $5$ independent groups, where no baselines were repeated in each group. Then, baselines within each group were averaged together, and the groups were cross-multiplied to form power spectra. This grouping method was used to reduce computational time, however upon closer examination it has been found that the initial grouping introduces an element of randomness into the final measurements --- more specifically, the power spectrum value fluctuates depending on how baselines are assigned into their initial groups. Our new approach removes this element of randomness at the cost of computational expense, as we now perform all baseline cross-products.}

\item{Finally, the last change from the \citetalias{ali_et_al2015} method is that our power spectrum points (previously computed as the mean of all bootstraps), are now computed as the power spectrum estimate resulting from not bootstrapping at all. More specifically, we compute one estimate without sampling, and this estimate is propagated through our signal loss computation (this estimate is $\widehat{\textbf{P}}_{x}$). The difference between taking the mean of the bootstrapped values and using the estimate from the no-bootstrapping case is small, but doing the latter ensures that we are forming results that reflect the estimate preferred by all our data.}

@@

The second null test possibility (error bars consistent with zero but not with thermal noise) is displayed by the $k$-values just outside the horizon ($k\sim\pm 0.1$\,$h$ Mpc$^{-1}$) for all three tests. This indicates an additive noise component that is increasing our errors. More specifically, although we expect each cross-multiplication that is used in power spectrum estimation to have independent noise, there is still the possibility of a noise-foreground coupling term that can introduce power. This is because cross-multiplications produce four additive terms --- a signal-squared term (where signal includes both foregrounds and EoR), two cross-terms between the signal and noise, and one noise-only term. When differencing two power spectra (each with their own four terms), we expect the signal-only term to subtract out for a ``passing" null test, and we expect the noise-only terms to be consistent with thermal noise. While the cross-terms have a mathematical expectation value of zero, in practice we are limited by our number of samples ($90$ cross-products for this analysis times $\sim$ $8$ independent LST samples). Combined with the fact that the foregrounds are so bright, these couplings can introduce extra variance that would vary with foreground strength. It is therefore not surprising that this effect is largest at $k$-values just outside the horizon, where we expect foregrounds to be brightest post-delay filtering.
